---
title: "Reproducible Research: Peer Assessment 1"
author: "Peter Willmott"
date: "July 28, 2018
output: 
  html_document:
    keep_md: true
---
# Reproducible Research: Peer Assessment 1

Project assumes the data, "activity.zip", is in the working directory

## Loading and preprocessing the data

Unzipping, reading and changing the data variable from 'factor' to 'date' + creating clean dataset

```{r load}
unzip(zipfile="activity.zip")
data <- read.csv("activity.csv")
data$date <- as.Date(data$date)
data_omit <- na.omit(data)
```

## What is mean total number of steps taken per day?
Calculate the total number of steps taken per day and show as Histogram

```{r total}
library(ggplot2)

data_aggr_sum <- aggregate(data_omit$steps ~ data_omit$date, FUN = "sum")
qplot(data_aggr_sum$`data_omit$steps`, bins = 15, xlab = "Steps / Day")
```

Calculate and report the mean and median of the total number of steps taken per day

```{r mean}
mean(data_aggr_sum$`data_omit$steps`)
median(data_aggr_sum$`data_omit$steps`)
```

## What is the average daily activity pattern?

Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r time_series}
library(lattice)

data_aggr_mean <- aggregate(data_omit$steps ~ data_omit$interval, FUN = "mean")
xyplot(data_aggr_mean$`data_omit$steps`~ data_aggr_mean$`data_omit$interval`, type="l", ylab="Steps (average)", xlab="Intervals (5 minutes")
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r max}
index <- which.max(data_aggr_mean$`data_omit$steps`)
data_aggr_mean[index,]
```

## Imputing missing values
There are a number of days/intervals where there are missing values. The presence of missing days may introduce bias into some calculations or summaries of the data.

Calculate and report the total number of missing values in the dataset.

```{r missing}
sum(is.na(data$steps))
```

Fill in all of the missing values in the dataset using the mean for that interval
The mean data is avalible from the previous question on average daily activity pattern.

```{r missing}
imputing <- data
for(i in 1:17568){
  if (is.na(imputing$steps[i])) {
    tochange <- data_aggr_mean$`data_omit$steps`[data_aggr_mean$`data_omit$interval` == imputing$interval[i]];
    imputing$steps[i] <- tochange;
  }
}
```

Make a histogram of the total number of steps taken each day

```{r total2}
data_aggr_sum_noNA <- aggregate(imputing$steps ~ imputing$date, FUN = "sum")
qplot(data_aggr_sum_noNA$`imputing$steps`, bins = 15, xlab = "Steps / Day")
```
Calculate and report the mean and median total number of steps taken per day. 

```{r mean2}
mean(data_aggr_sum_noNA$`imputing$steps`)
median(data_aggr_sum_noNA$`imputing$steps`)
```
The shape of the graph does not differ from the estimates from the first part of the asignment. Using the mean steps for a specific interval to impute the NA values causes little impact to the mean and median but do cause an increase in the number of steps taken per day for a specific interval, as more data has been added to the analysis.


## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day. 

```{r day}
dayweek <- imputing
dayweek$day <- weekdays(dayweek$date)
for(x in 1:17568){
  if(dayweek$day[x] == "Saturday" | dayweek$day[x] == "Sunday"){
    dayweek$levels[x] <- "weekend"
  }
  else {
    dayweek$levels[x] <- "weekday"
  }
}
```
Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r time_series}
data_aggr_mean_day <- aggregate(dayweek$steps ~ dayweek$interval + dayweek$levels, FUN = "mean")
xyplot(data_aggr_mean_day$`dayweek$steps` ~ data_aggr_mean_day$`dayweek$interval` | data_aggr_mean_day$`dayweek$levels`, layout = c(1,2), type="l", ylab="Steps (average)", xlab="Intervals (5 minutes", main = "Weekdays VS. Weekends")
```
### END!
